---
title: "Why Do Employees Leave?"
subtitle: "HR Attrition Analysis — IBM Dataset"
author: "Frenkli Paluku & Fatir Beqollari | Data Science, Faculty of Natural Sciences, Tirana"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    theme: flatly
    highlight: tango
    css: |
      @import url('https://fonts.googleapis.com/css2?family=Inter:wght@300;400;600;700&family=JetBrains+Mono&display=swap');

      body {
        font-family: 'Inter', sans-serif;
        background-color: #0f1117;
        color: #e2e8f0;
        font-size: 15px;
        line-height: 1.75;
      }

      h1, h2, h3 {
        font-family: 'Inter', sans-serif;
        font-weight: 700;
        color: #f8fafc;
        letter-spacing: -0.5px;
      }

      h1.title {
        font-size: 2.4em;
        color: #60a5fa;
        border-bottom: 3px solid #3b82f6;
        padding-bottom: 12px;
        margin-bottom: 6px;
      }

      h3.subtitle {
        color: #94a3b8;
        font-weight: 400;
        font-size: 1.1em;
        margin-top: 0;
      }

      h4.author {
        color: #60a5fa;
        font-size: 0.95em;
        font-weight: 600;
      }

      h4.date {
        color: #64748b;
        font-size: 0.85em;
        font-weight: 400;
      }

      /* Section headings */
      .section h1 { color: #60a5fa; border-left: 4px solid #3b82f6; padding-left: 14px; }
      .section h2 { color: #93c5fd; border-left: 3px solid #60a5fa; padding-left: 10px; }
      .section h3 { color: #bfdbfe; font-size: 1.05em; }

      /* Code blocks */
      pre, code {
        font-family: 'JetBrains Mono', monospace;
        font-size: 0.82em;
        background-color: #1e2433 !important;
        color: #a5f3fc !important;
        border: 1px solid #2d3748;
        border-radius: 8px;
      }

      pre {
        padding: 14px 18px;
        overflow-x: auto;
        box-shadow: 0 2px 8px rgba(0,0,0,0.4);
      }

      /* TOC */
      #TOC {
        background-color: #1a1f2e;
        border: 1px solid #2d3748;
        border-radius: 10px;
        padding: 16px;
        font-size: 0.88em;
      }

      #TOC a { color: #94a3b8; text-decoration: none; }
      #TOC a:hover { color: #60a5fa; }
      #TOC li.active > a { color: #60a5fa; font-weight: 600; }

      /* Tables */
      table {
        border-collapse: collapse;
        width: 100%;
        background-color: #1a1f2e;
        border-radius: 10px;
        overflow: hidden;
        margin: 16px 0;
        font-size: 0.9em;
      }

      th {
        background-color: #1e3a5f;
        color: #93c5fd;
        padding: 12px 16px;
        text-align: left;
        font-weight: 600;
        text-transform: uppercase;
        letter-spacing: 0.5px;
        font-size: 0.8em;
      }

      td {
        padding: 10px 16px;
        color: #cbd5e1;
        border-bottom: 1px solid #2d3748;
      }

      tr:last-child td { border-bottom: none; }
      tr:hover td { background-color: #1e2433; }

      /* Blockquotes as callout boxes */
      blockquote {
        background-color: #1e2d45;
        border-left: 4px solid #3b82f6;
        border-radius: 0 8px 8px 0;
        margin: 20px 0;
        padding: 14px 20px;
        color: #bfdbfe;
        font-size: 0.95em;
      }

      /* Links */
      a { color: #60a5fa; }
      a:hover { color: #93c5fd; }

      /* Main content area */
      .main-container { max-width: 960px; }

      /* Horizontal rules */
      hr { border-color: #2d3748; margin: 28px 0; }

      /* Strong text */
      strong { color: #93c5fd; }

      /* Plot container */
      img { border-radius: 10px; box-shadow: 0 4px 16px rgba(0,0,0,0.5); }
---

---

# Executive Summary

This project looks at **why employees leave their jobs**, using data from 1,470 IBM employees. We cleaned the data, explored it, and visualized the key patterns.

> **Bottom line in one sentence:** The biggest reason employees leave is **working overtime** — they leave at nearly 3× the rate of employees who don't.

**The 4 main findings:**

| Factor | Finding |
|---|---|
| Overtime | Overtime workers leave at **30.5%** vs. **10.4%** — the biggest gap we found |
| Age | Employees aged **18–25 leave at 35.8%** — double the company average |
| Department | **Sales (20.6%)** and **HR (19%)** are the most at-risk departments |
| Distance | Employees living far from work leave **more often** |

The overall attrition rate is **16.1%** — about 1 in 6 employees left.

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center"
)
```

# Setup

## Libraries

```{r}
library(dplyr)
library(readr)
library(ggplot2)
library(skimr)
library(corrplot)
library(gridExtra)
library(RColorBrewer)
```

## Custom Plot Theme

We built a custom dark theme that matches the report's style. Every chart in this project uses it.

```{r}
# Dark palette — two main colors for attrition groups
pal_main  <- c("#3b82f6", "#ef4444")   # blue = stayed, red = left
pal_multi <- c("#3b82f6","#60a5fa","#93c5fd","#bfdbfe","#dbeafe")

my_theme <- theme_minimal(base_size = 13) +
  theme(
    # Background
    plot.background    = element_rect(fill = "#0f1117", color = NA),
    panel.background   = element_rect(fill = "#1a1f2e", color = NA),
    panel.grid.major   = element_line(color = "#2d3748", linewidth = 0.4),
    panel.grid.minor   = element_blank(),

    # Text
    plot.title         = element_text(size = 15, face = "bold",
                                      color = "#f8fafc", margin = margin(b = 4)),
    plot.subtitle      = element_text(size = 11, color = "#94a3b8",
                                      margin = margin(b = 10)),
    axis.title         = element_text(size = 11, color = "#94a3b8"),
    axis.text          = element_text(size = 10, color = "#cbd5e1"),

    # Legend
    legend.background  = element_rect(fill = "#1a1f2e", color = NA),
    legend.text        = element_text(color = "#cbd5e1"),
    legend.title       = element_text(color = "#94a3b8", face = "bold"),

    # Borders
    plot.margin        = margin(16, 16, 16, 16)
  )
```

---

# Loading the Data

```{r}
HR_Employee_Attrition <- read_csv("HR-Employee-Attrition.csv")
```

## First Look

```{r}
skim(HR_Employee_Attrition)
```

**What we see:**

- **1,470 employees**, **35 columns**
- **No missing values** anywhere — the data is clean from the start
- 3 columns (`EmployeeCount`, `StandardHours`, `Over18`) are the **same value for every row** — they tell us nothing, so we'll drop them

---

# Data Cleaning

## Step 1 — Pick the Useful Columns

We keep 8 columns that are most relevant to attrition. The other 27 are either IDs, constants, or outside our focus.

```{r}
HR_Kept <- HR_Employee_Attrition %>%
  select(Age, Attrition, BusinessTravel, DailyRate,
         Department, DistanceFromHome, Education, OverTime)

head(HR_Kept)
```

## Step 2 — Rename for Clarity

Shorter, cleaner names make the code easier to read.

```{r}
HR_Renamed <- HR_Kept %>%
  rename(
    Left     = Attrition,
    Pay      = DailyRate,
    Distance = DistanceFromHome,
    Travel   = BusinessTravel,
    Overtime = OverTime
  )

head(HR_Renamed)
```

## Step 3 — Check for Missing Values

```{r}
colSums(is.na(HR_Renamed))
```

All zeros. **No missing data** — nothing to fix here.

## Step 4 — Remove Duplicates

```{r}
HR_NoDupes <- HR_Renamed %>% distinct()
any(duplicated(HR_NoDupes))
```

`FALSE` means no duplicates. All 1,470 rows are unique.

## Step 5 — Convert Yes/No to 1/0

Computers work better with numbers. We turn `Left` and `Overtime` from text into 1 (Yes) or 0 (No).

```{r}
HR_Numericals <- HR_NoDupes %>%
  mutate(
    Left     = ifelse(Left == "Yes", 1, 0),
    Overtime = ifelse(Overtime == "Yes", 1, 0)
  )

head(HR_Numericals)
```

---

# Feature Engineering

We create **4 new columns** to make patterns easier to see in charts.

```{r}
HR_Final <- HR_Numericals %>%
  mutate(
    # Group age into life stages
    AgeGroup = cut(Age,
      breaks = c(18, 25, 35, 45, 55, 60),
      labels = c("18-25", "26-35", "36-45", "46-55", "56-60"),
      include.lowest = TRUE),

    # Group commute distance into bands
    DistanceGroup = cut(Distance,
      breaks = c(0, 5, 10, 20, 30),
      labels = c("Very Close", "Close", "Moderate", "Far"),
      include.lowest = TRUE),

    # Turn education codes into readable labels (ordered so charts sort correctly)
    EducationLevel = factor(
      case_when(
        Education == 1 ~ "Below College",
        Education == 2 ~ "College",
        Education == 3 ~ "Bachelor",
        Education == 4 ~ "Master",
        Education == 5 ~ "Doctor"
      ),
      levels  = c("Below College","College","Bachelor","Master","Doctor"),
      ordered = TRUE),

    # Standardize pay so it can be compared in correlation (mean=0, sd=1)
    PayScaled = as.numeric(scale(Pay))
  )

head(HR_Final)
```

**What each new column does:**

- `AgeGroup` — splits age into 5 career stages for cleaner bar charts
- `DistanceGroup` — turns raw distance numbers into 4 meaningful bands
- `EducationLevel` — readable labels, sorted correctly (not alphabetically)
- `PayScaled` — standardized pay used in the correlation analysis

---

# Exploratory Analysis

## Age Range

```{r}
range(HR_Final$Age)
```

Employees are between **18 and 60 years old** — the full working-age range.

## How Many Employees Left?

```{r}
table(HR_Final$Left)
prop.table(table(HR_Final$Left)) * 100
```

**237 employees left (16.1%)**, 1,233 stayed (83.9%).

Because leavers are the minority, we'll always use **percentages**, not raw counts. Otherwise "stayed" would always dominate the charts and hide the real patterns.

---

# Summary Statistics

## Overall Attrition Rate

```{r}
mean(HR_Final$Left) * 100
```

**16.1%** is our company baseline. Any group above this is more at risk than average.

## By Overtime

```{r}
HR_Final %>%
  group_by(Overtime) %>%
  summarise(AttritionRate = round(mean(Left) * 100, 1))
```

> **30.5% vs 10.4%** — overtime workers leave at nearly 3 times the rate. This is the biggest difference in the entire dataset.

## By Department

```{r}
HR_Final %>%
  group_by(Department) %>%
  summarise(AttritionRate = round(mean(Left) * 100, 1))
```

> **Sales (20.6%) and HR (19%)** are above the 16.1% average. R&D (13.8%) is the most stable.

## By Age Group

```{r}
HR_Final %>%
  group_by(AgeGroup) %>%
  summarise(AttritionRate = round(mean(Left) * 100, 1))
```

> The **18–25 group leaves at 35.8%** — more than double the company average. Attrition falls sharply after age 35.

---

# Visualizations

## Attrition by Age Group

```{r fig.width=8, fig.height=5}
p1 <- HR_Final %>%
  group_by(AgeGroup) %>%
  summarise(AttritionRate = mean(Left) * 100) %>%
  ggplot(aes(x = AgeGroup, y = AttritionRate, fill = AgeGroup)) +
  geom_col() +
  geom_hline(yintercept = 16.1, linetype = "dashed",
             color = "#f59e0b", linewidth = 0.8) +
  annotate("text", x = 4.6, y = 17.5, label = "Avg: 16.1%",
           color = "#f59e0b", size = 3.5, fontface = "bold") +
  scale_fill_manual(values = pal_multi) +
  labs(title    = "Attrition Rate by Age Group",
       subtitle = "Dashed line = company average (16.1%)",
       x = "Age Group", y = "Attrition Rate (%)") +
  my_theme + theme(legend.position = "none")

p1
```

**The 18–25 group is at 35.8%** — way above the average. After 35, most employees stabilize. Younger employees are harder to retain — likely due to fewer ties, more career exploration, and less seniority.

## Attrition by Overtime

```{r fig.width=7, fig.height=5}
p2 <- HR_Final %>%
  group_by(Overtime) %>%
  summarise(AttritionRate = mean(Left) * 100) %>%
  ggplot(aes(x = factor(Overtime, labels = c("No Overtime","Overtime")),
             y = AttritionRate,
             fill = factor(Overtime))) +
  geom_col() +
  geom_text(aes(label = paste0(round(AttritionRate, 1), "%")),
            vjust = -0.5, color = "#f8fafc", size = 5, fontface = "bold") +
  scale_fill_manual(values = c("#3b82f6","#ef4444")) +
  labs(title    = "Attrition Rate by Overtime",
       subtitle = "Overtime nearly triples the chance of leaving",
       x = NULL, y = "Attrition Rate (%)") +
  my_theme + theme(legend.position = "none")

p2
```

**30.5% vs 10.4%.** This is the single clearest pattern in the data — working overtime is the strongest predictor of attrition.

## Attrition by Department

```{r fig.width=8, fig.height=5}
p3 <- HR_Final %>%
  group_by(Department) %>%
  summarise(AttritionRate = mean(Left) * 100) %>%
  ggplot(aes(x = reorder(Department, AttritionRate),
             y = AttritionRate, fill = AttritionRate)) +
  geom_col() +
  geom_hline(yintercept = 16.1, linetype = "dashed",
             color = "#f59e0b", linewidth = 0.8) +
  coord_flip() +
  scale_fill_gradient(low = "#3b82f6", high = "#ef4444") +
  labs(title    = "Attrition Rate by Department",
       subtitle = "Sorted by attrition — dashed = company average",
       x = NULL, y = "Attrition Rate (%)") +
  my_theme + theme(legend.position = "none")

p3
```

**Sales is the most at-risk department at 20.6%**, followed closely by HR. R&D at 13.8% is the most stable — likely due to more specialized roles and clearer career ladders.

## Attrition by Distance from Home

```{r fig.width=8, fig.height=5}
p4 <- HR_Final %>%
  group_by(DistanceGroup) %>%
  summarise(AttritionRate = mean(Left) * 100) %>%
  ggplot(aes(x = DistanceGroup, y = AttritionRate, fill = DistanceGroup)) +
  geom_col() +
  scale_fill_manual(values = pal_multi) +
  labs(title    = "Attrition Rate by Distance from Home",
       subtitle = "Longer commutes = higher chance of leaving",
       x = "Distance Band", y = "Attrition Rate (%)") +
  my_theme + theme(legend.position = "none")

p4
```

Employees living far from work leave more. The effect is real but smaller than overtime or age — commute alone rarely causes attrition, but it adds up alongside other pressures.

## Combined View

```{r fig.width=13, fig.height=10}
grid.arrange(p1, p2, p3, p4, ncol = 2,
             top = grid::textGrob("Four Key Drivers of Employee Attrition",
                                  gp = grid::gpar(fontsize = 16,
                                                  fontface = "bold",
                                                  col = "#f8fafc")))
```

---

# Advanced Visualizations

## Heatmap: Department × Overtime

This chart shows what happens when two risk factors combine.

```{r fig.width=9, fig.height=5}
HR_Final %>%
  group_by(Department, Overtime) %>%
  summarise(AttritionRate = mean(Left) * 100, .groups = "drop") %>%
  ggplot(aes(x = Department,
             y = factor(Overtime, labels = c("No Overtime","Overtime")),
             fill = AttritionRate)) +
  geom_tile(color = "#0f1117", linewidth = 1.5) +
  geom_text(aes(label = paste0(round(AttritionRate, 1), "%")),
            size = 6, fontface = "bold", color = "white") +
  scale_fill_gradient(low = "#1e3a5f", high = "#dc2626") +
  labs(title    = "Attrition Heatmap: Department × Overtime",
       subtitle = "Darkest red = highest risk group",
       x = NULL, y = NULL, fill = "Attrition %") +
  my_theme
```

**Sales + Overtime is the highest-risk combination** in the company. This is something a simple bar chart can't reveal — you need to look at two factors together. An employee in Sales who works overtime is at serious risk of leaving.

---

# Boxplots — Distributions, Not Just Averages

A boxplot shows the **full spread** of values, not just the average. The line in the middle is the median.

## Pay by Attrition

```{r fig.width=7, fig.height=5}
ggplot(HR_Final,
       aes(x = factor(Left, labels = c("Stayed","Left")),
           y = Pay, fill = factor(Left))) +
  geom_boxplot(outlier.color = "#f59e0b", outlier.size = 1.5) +
  scale_fill_manual(values = c("#3b82f6","#ef4444")) +
  labs(title    = "Daily Pay by Attrition Status",
       subtitle = "Similar medians — pay alone doesn't explain leaving",
       x = NULL, y = "Daily Pay Rate") +
  my_theme + theme(legend.position = "none")
```

Both groups have nearly identical pay distributions. **Pay rate is not a meaningful driver of attrition** in this dataset.

## Age by Attrition

```{r fig.width=7, fig.height=5}
ggplot(HR_Final,
       aes(x = factor(Left, labels = c("Stayed","Left")),
           y = Age, fill = factor(Left))) +
  geom_boxplot(outlier.color = "#f59e0b", outlier.size = 1.5) +
  scale_fill_manual(values = c("#3b82f6","#ef4444")) +
  labs(title    = "Age Distribution by Attrition Status",
       subtitle = "Employees who left are notably younger",
       x = NULL, y = "Age") +
  my_theme + theme(legend.position = "none")
```

The median age is clearly **lower for employees who left**. This confirms the age group bar chart — younger employees leave more.

## Distance by Attrition

```{r fig.width=7, fig.height=5}
ggplot(HR_Final,
       aes(x = factor(Left, labels = c("Stayed","Left")),
           y = Distance, fill = factor(Left))) +
  geom_boxplot(outlier.color = "#f59e0b", outlier.size = 1.5) +
  scale_fill_manual(values = c("#3b82f6","#ef4444")) +
  labs(title    = "Commute Distance by Attrition Status",
       subtitle = "Employees who left lived slightly further away",
       x = NULL, y = "Distance from Home") +
  my_theme + theme(legend.position = "none")
```

The difference is modest — commute contributes to attrition but isn't the main cause by itself.

---

# Pay Distribution — Histogram

```{r fig.width=8, fig.height=5}
ggplot(HR_Final, aes(x = Pay)) +
  geom_histogram(bins = 30, fill = "#3b82f6", color = "#0f1117", alpha = 0.85) +
  labs(title    = "Distribution of Daily Pay",
       subtitle = "Spread evenly — no salary cluster stands out",
       x = "Daily Pay Rate", y = "Number of Employees") +
  my_theme
```

Pay is spread fairly evenly across all salary levels. There's no obvious cluster, which is part of why pay doesn't separate leavers from stayers.

---

# Scatterplot — Age vs Pay

```{r fig.width=9, fig.height=6}
ggplot(HR_Final,
       aes(x = Age, y = Pay,
           color = factor(Left, labels = c("Stayed","Left")))) +
  geom_jitter(size = 1.6, alpha = 0.45, width = 0.5) +
  scale_color_manual(values = c("#3b82f6","#ef4444")) +
  labs(title    = "Age vs Daily Pay — by Attrition Status",
       subtitle = "Red and blue are mixed together — no clear separation",
       x = "Age", y = "Daily Pay Rate", color = "Status") +
  my_theme
```

The red dots (left) and blue dots (stayed) are mixed throughout the entire chart. **No region is dominated by one group.** Age and pay together do not predict attrition well — the real driver is overtime.

---

# Correlation Analysis

We measure how strongly each variable is mathematically linked to attrition. `PayScaled` is the standardized version of Pay, included so it can be compared fairly.

```{r}
cor_data <- HR_Final %>%
  select(Age, PayScaled, Distance, Overtime, Left)

cor_matrix <- cor(cor_data)
round(cor_matrix, 3)
```

## Correlation Heatmap

```{r fig.width=8, fig.height=6}
corrplot(cor_matrix,
         method      = "color",
         type        = "upper",
         addCoef.col = "#f8fafc",
         number.cex  = 0.85,
         tl.cex      = 0.9,
         tl.col      = "#e2e8f0",
         col         = colorRampPalette(c("#1e3a5f","#0f1117","#7f1d1d"))(200),
         bg          = "#1a1f2e",
         mar         = c(0, 0, 1, 0))
```

**Reading the `Left` column:**

| Variable | Correlation with Attrition | Meaning |
|---|---|---|
| Overtime | **+0.25** | Strongest link — overtime → more likely to leave |
| Age | **–0.16** | Older → less likely to leave |
| Distance | +0.08 | Slight effect — farther = slightly more likely to leave |
| PayScaled | –0.06 | Almost no link — pay barely matters |

These numbers confirm everything we saw in the charts.

---

# Unique Analysis: Attrition Risk Score

*This section calculates a simple risk score for each employee — something that goes beyond individual charts and combines factors into a single number.*

We assign points based on the three most impactful factors: **Overtime, Age Group, and Department**. An employee with all three risk factors gets the highest score.

```{r fig.width=9, fig.height=5}
# Assign risk points to each employee
HR_Risk <- HR_Final %>%
  mutate(
    # +3 for overtime (biggest factor)
    RiskOT   = ifelse(Overtime == 1, 3, 0),

    # +2 for being in the youngest group, +1 for 26-35
    RiskAge  = case_when(
      AgeGroup == "18-25" ~ 2,
      AgeGroup == "26-35" ~ 1,
      TRUE ~ 0
    ),

    # +1 for being in Sales or HR (above-average departments)
    RiskDept = ifelse(Department %in% c("Sales","Human Resources"), 1, 0),

    # Total risk score 0-6
    RiskScore = RiskOT + RiskAge + RiskDept
  )

# Actual attrition rate for each risk score
risk_summary <- HR_Risk %>%
  group_by(RiskScore) %>%
  summarise(
    Employees     = n(),
    AttritionRate = round(mean(Left) * 100, 1)
  )

risk_summary

# Plot
ggplot(risk_summary,
       aes(x = factor(RiskScore), y = AttritionRate, fill = AttritionRate)) +
  geom_col() +
  geom_text(aes(label = paste0(AttritionRate, "%\n(n=", Employees, ")")),
            vjust = -0.4, color = "#f8fafc", size = 3.8, fontface = "bold") +
  scale_fill_gradient(low = "#1e3a5f", high = "#dc2626") +
  labs(
    title    = "Attrition Rate by Composite Risk Score",
    subtitle = "Score combines Overtime (3pts) + Age Risk (2pts) + Department Risk (1pt)",
    x        = "Risk Score (0 = Low Risk, 6 = Highest Risk)",
    y        = "Actual Attrition Rate (%)"
  ) +
  my_theme + theme(legend.position = "none")
```

**This is the key takeaway from the whole project.** Employees with a risk score of 0 (no risk factors) leave at just ~8%. Employees with a score of 5–6 (all risk factors) can leave at over 50%. This shows that **risk factors multiply each other** — they don't just add up independently.

---

# Conclusion

We analyzed 1,470 employee records to understand what drives attrition. Here's what the data shows:

| # | Finding | In plain words |
|---|---|---|
| 1 | **Overtime → 30.5% attrition** | Overworked employees leave. Simple. |
| 2 | **Age 18–25 → 35.8% attrition** | Young employees are hard to keep. |
| 3 | **Sales & HR above average** | These departments need more attention. |
| 4 | **Far distance → higher attrition** | Long commutes add up over time. |
| 5 | **Pay rate → almost no effect** | Paying more daily doesn't fix the core issues. |

The preprocessing steps (removing constants, checking for missing values, renaming, encoding, and engineering features) were essential — they turned a messy 35-column dataset into something we could actually analyze and visualize clearly.

---

# Recommendations

Based on the data, here is what the company should prioritize:

1. **Cut overtime** — especially in Sales. It's the #1 predictor and the most directly fixable.
2. **Support young employees** — mentorship, career growth plans, and clear promotion paths help the 18–25 group stay.
3. **Investigate Sales and HR** — exit interview data from these departments would reveal what's not in the numbers.
4. **Offer flexibility for long commuters** — remote or hybrid work options reduce commute stress before it becomes a reason to quit.

---

*Frenkli Paluku & Fatir Beqollari — Data Science, Faculty of Natural Sciences, University of Tirana*
