---
title: "Data Preprocessing and Visualization : Assigment 1 "
author: Frenkli Paluku
output:
  html_document:
    df_print: paged
---
**Accepted by:**  
**Lab Lecturer:** Arnisa Sokoli  
**Lecturer:** Lule Basha
```{r}
library(tidyverse)
ds = msleep
```


## Task 1: Logical Duplicate Handling
1. Identify if there are duplicate entries in the name column.

```{r}
#We use the any function to see if the column NAME has any duplicates
any(duplicated(ds$name))
#The function returned FALSE so it means the column has no duplicates
```

2. Remove a row only if both the name and the genus are identical.
```{r}
#We use nrow function to compare the number of rows before and after the distinct of name == genus if any occured
nrow(ds)
nrow(distinct(ds, name, genus, .keep_all = TRUE))
#We can see that number is still the same so no row removed
```
3. Rename the column vore to diet_category and conservation to protection_status
```{r}
#We use the Rename function to change the name of the column vore and conservation
ds <- rename(ds , diet_category = vore, protection_status = conservation)
colnames(ds)
#We can now see that the columns we wanted have had their name changed
```
## Task 2: Advanced Imputation
1. Calculate the mean of sleep_cycle grouped by the order column
```{r}
#We first use GROUPBY function to group the data by their ORDER column and then we use use the summarise of the mean sleep_cycle without NA values
ds %>%
  group_by(order) %>%
  summarise(mean_sleep_cycle = mean(sleep_cycle, na.rm = TRUE))
#Some of the results are NaN becuase no animal in that ORDER had a value in the sleep
```
2. Replace the NA values in sleep_cycle with the average of their respective taxonomic order
```{r}
#We do the same group_by as above but now we use mutate to replace any value that is Na to its ORDER average else we keep it as it was
#Here we also use a counter to see the before and after of how many Na values we have and will have
sum(is.na(ds$sleep_cycle))
ds <- ds %>%
  group_by(order) %>%
    mutate(sleep_cycle = ifelse(is.na(sleep_cycle), mean(sleep_cycle, na.rm = TRUE), sleep_cycle))
#After
sum(is.na(ds$sleep_cycle))
#We can see that we replaced 39 Na values with their VALUE average
#We couldnt replace all becuase some classes have no sleep_cycle value at any instance
```
3. If an entire order consists of NA values, fill those remaining gaps with the median of the
entire sleep_cycle variable.
```{r}
#First we get the median of the sleep_cycle column
med <- median(ds$sleep_cycle , na.rm = TRUE)
print(med)
#Then we replace the Na values in that column with the median
ds$sleep_cycle[is.na(ds$sleep_cycle)] <- med
sum(is.na(ds$sleep_cycle))
#Now we can see that we replaced the last 12 remained Na values
```
## Task 3: Outlier Detection via Log-Transformation
1. Apply a log10 transformation to bodywt and store it in a new column log_weight. 2. Using
the 1.5 * IQR rule on the log_weight column, identify the upper and lower bounds for outliers.
```{r}
#We create another column that is log10 of the bodywt
ds$log_weight <- log10(ds$bodywt)
#We find the Q1 and Q3 using the QUANTILE function
Q1 <- quantile(ds$log_weight, 0.25, na.rm = TRUE)
Q3 <- quantile(ds$log_weight, 0.75, na.rm = TRUE)
#We find the IQR using the IQR function
IQR <- IQR(ds$log_weight, na.rm = TRUE)
#We find the lower & upper bond using simple math
lower_bound <- Q1 - 1.5 * IQR
upper_bound <- Q3 + 1.5 * IQR
print(paste("The value for Q1 is", Q1))
print(paste("The value for Q3 is", Q3))
print(paste("The value for IQR is", IQR))
print(paste("The value for lower bound is", lower_bound))
print(paste("The value for upper bound is", upper_bound))
```
3. Create a logical column is_outlier that flags rows falling outside these bounds.
```{r}
#We create another new column that has the boolean values if a values is or not an outlier
ds$is_outlier <- ds$log_weight < lower_bound | ds$log_weight > upper_bound
print(ds$is_outlier)
#As we can see no value is an outlier
```
Task 4: Categorical and Numerical Transformation
1. Create a new column animal_class which contains:
• "Megafauna Herbivore": if vore is "herbi" AND bodywt > 200.
• "Standard Herbivore": if vore is "herbi" AND bodywt <= 200.
• "Non-Vegetarian": if vore is "omni" OR "carni".
• "Unclassified": for any other case or NA
```{r}
#We created a new column that includes all the condtions above using ifelse for each
ds$animal_class <- ifelse(ds$diet_category == "herbi" & ds$bodywt > 200, "Megafauna Herbivore",
                   ifelse(ds$diet_category == "herbi" & ds$bodywt <= 200, "Standard Herbivore",
                   ifelse(ds$diet_category == "omni" | ds$diet_category == "carni", "Non-Vegetarian",
                          "Unclassified")))
print(ds$animal_class)
#The results are 4 varies as expected
```
2. Normalize the weight column to a scale of 0 to 1 (Min-Max Scaling) and store the values in a new
column
```{r}
#We use the Normalization formula to normalize the values to 0-1 scale
ds$bodywt_scaled <- (ds$bodywt - min(ds$bodywt, na.rm = TRUE)) / (max(ds$bodywt, na.rm = TRUE) - min(ds$bodywt, na.rm = TRUE))
print(round(ds$bodywt_scaled , 3))
#The results are as expected all the values expand on a 0-1 range
```
## Task 5: The Pipeline Challenge
Write a single continuous dplyr pipe that:
1. Filters for animals with a brainwt greater than the average.
2. Computes sleep_efficiency as (sleep_total / 24) * 100.
3. Selects only name, sleep_efficiency, and protection_status.
4. Arranges the results by sleep_efficiency in descending order.
```{r}
#We use the FILTER function to filter the data greater than the mean , We use the MUTATE to get a new column that has (sleep_total / 24) * 100 values in it , We use the SELECT to only select the three columns we need and , We use arrange to arrange the results in desending order by sleep
ds %>%
  filter(brainwt > mean(brainwt, na.rm = TRUE)) %>%
  mutate(sleep_efficiency = (sleep_total / 24) * 100) %>%
  select(name, sleep_efficiency, protection_status) %>%
  arrange(desc(sleep_efficiency))
```




  







